---
title: "Study on stress-adaptation and visuospatial social memory: Power analysis and sample size determination for Mozambique replication study"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(extrafont)
library(faux)
library(dplyr)
library(lme4)
library(lmerTest)
library(ggplot2)
library(tidyr)
library(broom.mixed)
library(purrr)
library(reshape2)
library(data.table)

### GRAPH THEMES ###

#fonts
quartzFonts(avenir = c("Avenir Book", "Avenir Black", "Avenir Book Oblique", "Avenir Black Oblique"))

#theme
header_size = 10
axis_size = 10

#theme for plots
avenir_theme = theme(text=element_text(size=header_size,family='avenir'),
                     axis.text.x = element_text(color = 'black', size = axis_size, vjust = 1),
                     axis.text.y = element_text(color = 'black', size = axis_size),
                     axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), face = "bold"),
                     axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 0, l = 0), face = "bold"),
                     panel.background = element_blank(),
                     panel.grid.major.x = element_line(color = '#e7e7e7'),
                     panel.grid.major.y = element_line(color = '#e7e7e7'),
                     legend.key = element_blank(),
                     legend.title = element_text(color = 'black', size = axis_size, face = "bold"),
                     plot.title = element_text(hjust = 0.5, face = "bold"))

#set the font
par(family = 'avenir')

```

#### Gabriele Paone, Arran J Davis, Emma Cohen
##### Institute of Human Sciences, University of Oxford
##### arran.davis@anthro.ox.ac.uk | davis.arran@gmail.com
##### 29 June 2022

## Introduction
We hypothesize that children from higher-stress areas of Gorongosa National Park in Mozambique will score higher in working and short-term memory tasks that use social stimuli than will children from a lower-stress areas of the park. 

## Data simulation
The outcome variables are performance in two tasks measuring: (1) visuospatial social short-term memory and (2) visuospatial social working memory.

The study will include participants from different backgrounds in Gorongosa National Park: one group with a lower-income background and one group with a higher-income background.

To estimate social short-term memory and social working memory scores for this study, we used as a reference the estimated mean differences from regression models used to analyse results from our previous study of the same variables in different populations from Napoli, Italy. We found mean differences between a lower-income group (Scampia neighbourhood) and a relatively higher income group (Pozzuoli neighbourhood) of 0.841 (_SE_ 0.131) points for social short-term memory and 0.499 (_SE_ = 0.156) for social short-term working memory. This led to the following data simulation. 

``` {r data simulation, out.width="100%", echo=FALSE}

### CREATE DATA ###

#create the neighbourhood variable 
between = list(neighbourhood = c(Scampia = "Scampia", 
                                 Pozzuoli = "Pozzuoli "))

#create the test variable
within = list(test = c("Social short-term memory",
                       "Social working memory"))

within_short = list(test = c("social-short-term",
                             "social-working"))

#create scores for each neighbourhood for visuospatial social short-term memory
scampia_vss_stm = 1.260 + 0.841 + (8 * 0.479)
pozzuoli_vss_stm = 1.260  + (8 * 0.479)

#create scores for each neighbourhood for visuospatial social working memory
scampia_vss_wm = 0.285 + 0.499 + (8 * 0.489)
pozzuoli_vss_wm = 0.285 + (8 * 0.489)

#create the mean scores for each neighbourhood on each memory test 
hood_means = list(Scampia = c(vss_stm = scampia_vss_stm, 
                              vss_wm = scampia_vss_wm),
                  Pozzuoli = c(vss_stm = pozzuoli_vss_stm,
                               vss_wm = pozzuoli_vss_wm))

#create the standard deviation scores for each neighbourhood on each memory test 
hood_sds = list(Scampia = c(vss_stm = 1, 
                            vss_wm = 1),
                Pozzuoli = c(vss_stm = 1,
                             vss_wm = 0.75))

#set the correlation of test score results for each neighbourhood
hood_cors = list(Scampia = .6, Pozzuoli = .6)

#create the dataframe
dat = sim_design(within_short, between, n = 50, 
                 mu = hood_means, sd = hood_sds, r = hood_cors,
                 empirical = FALSE, plot = FALSE)

#make the data long format
dat_long = melt(setDT(dat), id.vars = c("id","neighbourhood"), variable.name = "test")

#ensure no values are below 2 or above 8
dat_long$value = ifelse(dat_long$value < 2, 2,
                        ifelse(dat_long$value > 8, 8, dat_long$value))

#set contrasts of neighbourhood so that "Pozzuoli" is the reference
contrasts(dat_long$neighbourhood) = contr.treatment(2, base = 2)

#round the test scores to a whole number
dat_long$value = round(dat_long$value, 0)

### ### ###

#load plot themes
source("plot_theme.R")

#plot the data
test_labels = c("Social short-term memory",
                "Social working memory")

hood_colors = c("#999999", "#009E73", "#F0E442")

scores_by_hood = ggplot(dat_long, aes(x = test, y = value)) + 
                  geom_violin(aes(fill = neighbourhood), trim = TRUE, position = position_dodge(0.9)) +
                  geom_boxplot(aes(fill = neighbourhood), width = 0.15, position = position_dodge(0.9)) +
                  scale_x_discrete(labels = test_labels) +
                  scale_y_continuous(limits = c(0,8), breaks = c(seq(0,8,1))) +
                  scale_fill_manual(values = hood_colors, labels = c("Higher-stress", "Lower-stress")) +
                  labs(fill = "Environment type") +
                  ylab("Test score") +
                  xlab("Test type") +
                  aes(ymin = 0) +
                  avenir_theme

scores_by_hood

```

## Power analyses 

Power analyses were carried out using the `faux` package in R (DeBruine, 2021). Using the dataset described above as a reference, 25,000 simulated sample datasets were created for each of the two memory test types. This was done through sampling different simulated populations of test scores (`test_score`) that varied according to environment type (`environment`), sample size (15, 20, 25, 30, 35) and the effect of the higher-stress environment on test scores, which differed for social short-term memory scores (0.2, 0.4, 0.841, 1, 1.2) and social working memory scores (0.25, 0.499, 0.75,  1, 1.25).

The following model was run on each of the 25,000 simulated sample datasets; treatment contrasts where used for the `environment` variable, with the lower-stress environment being the baseline or reference category.

```
lm(test_score ~ environment, data = data)
```

The power analyses reported below were calculated as the frequency of the 25,000 _p_-values that were less than 0.05 for the _b_-coefficient that compared test scores from the higher-stress environment to test scores from the lower-stress environment.

### Social short-term memory

Below is the power analysis result with social short-term memory as the outcome. Results suggest that 25 participants per environment would give sufficient power (0.81) to detect mean differences of at least 0.841 (the effect observed in our previous study) between the higher and lower stress environments.

```{r load and process power analysis data for social short-term memory, fig.show="hold", echo=FALSE}

#load the simulation data
simulations = read.csv("../data/study2_replication_social_memory_data_simulations.csv")

#subset the data to each outcome type
stm_sims = subset(simulations, simulations$outcome == "Social short-term memory")
wm_sims = subset(simulations, simulations$outcome == "Social working memory")

#create dataset to plot power analysis for neighbourhood main effects (both high-stress neighbourhoods were estimated to have the same effect)
simulation_results_neighbourhood_stm = filter(stm_sims, term == "neighbourhood1") %>%
                                              group_by(high_stress_soc_stm_effect, participant_n) %>% 
                                              summarise(power = mean(p.value < .05), .groups = "drop")


simulation_results_neighbourhood_wm = filter(wm_sims, term == "neighbourhood1") %>%
                                           group_by(high_stress_soc_wm_effect, participant_n) %>% 
                                           summarise(power = mean(p.value < .05), .groups = "drop")

### ### ###

#plot power analyse for neighbourhood effect on social short-term memory
stm_pa = ggplot(aes(as.character(high_stress_soc_stm_effect), participant_n, fill = power), 
                data = simulation_results_neighbourhood_stm) +
          geom_tile() +
          geom_text(aes(label = sprintf("%.2f", power)), color = "black", size = 5) +
          scale_y_continuous(breaks = c(15, 20, 25, 30, 35)) +
          scale_fill_viridis_c(name = "Power",
                               limits = c(0, 1), 
                               breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                               labels = c(0, 0.2, 0.4, 0.6, 0.8, 1)) +
          xlab("High-stress environment effect on social short-term memory") + 
          ylab("Participant sample size (per environment)") +
          avenir_theme

stm_pa

```

### Social working memory

Below is the power analysis result with social short-term memory as the outcome. Results suggest that 25 participants per environment would give sufficient power (0.80) to detect mean differences of at least 0.75 between the higher and lower stress environments.

```{r load and process power analysis data for social working memory, fig.show="hold", echo=FALSE}

#plot power analyse for neighbourhood effect on social working memory
wm_pa = ggplot(aes(as.character(high_stress_soc_wm_effect), participant_n, fill = power), 
               data = simulation_results_neighbourhood_wm) +
           geom_tile() +
           geom_text(aes(label = sprintf("%.2f", power)), color = "black", size = 5) +
           scale_y_continuous(breaks = c(15, 20, 25, 30, 35)) +
           scale_fill_viridis_c(name = "Power",
                                limits = c(0, 1), 
                                breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                                labels = c(0, 0.2, 0.4, 0.6, 0.8, 1)) +
           xlab("High-stress environment effect on social working memory") + 
           ylab("Participant sample size (per neighbourhood)") +
           avenir_theme

wm_pa

```

## Citations

DeBruine L (2021). _faux: Simulation for Factorial Designs_. doi: 10.5281/zenodo.2669586, R package version 1.1.0, https://debruine.github.io/faux/. 
